---
title: "Create and run an analysis profile"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Create and run an analysis profile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Start from a package example

Do not copy a historical analysis folder with results. Copy one reviewed YAML
and the standalone launcher files from the installed package:

```{r locate-examples}
examples <- system.file(
  "extdata", "vignette_examples", package = "hotnetR2", mustWork = TRUE
)
list.files(examples)
```

From the project root:

```bash
mkdir analysis_E092_Gastric_GEjunc
cp "$(Rscript --vanilla -e 'cat(system.file("extdata", "vignette_examples", package="hotnetR2"))')/analysis_aracne.yaml" \
   analysis_E092_Gastric_GEjunc/analysis.yaml
cp "$(Rscript --vanilla -e 'cat(system.file("extdata", "vignette_examples", package="hotnetR2"))')/run_analysis.sh" \
   analysis_E092_Gastric_GEjunc/
cp "$(Rscript --vanilla -e 'cat(system.file("extdata", "vignette_examples", package="hotnetR2"))')/run_analysis.R" \
   analysis_E092_Gastric_GEjunc/
chmod +x analysis_E092_Gastric_GEjunc/run_analysis.sh
```

For STRINGdb, substitute `analysis_stringdb.yaml`. Change the folder and
`analysis.name` together. The examples contain placeholder tool/input paths and
will not pass preflight until edited.

## Directory boundary

An analysis folder contains configuration and outputs only for that profile:

```text
analysis_name/
|-- analysis.yaml
|-- run_analysis.sh
|-- run_analysis.R
|-- metadata/
|-- logs/
|-- LDAK/
`-- hHotnet/
```

Shared GWAS, LD and caches normally live in the project root and are referenced
by paths relative to `project.root`. Never symlink a writable result directory
from an older analysis.

## Important YAML sections

| Section | Purpose |
|---|---|
| `analysis` | Unique name, label and analysis seed. |
| `project` | Root against which input paths resolve. |
| `gwas` | Processed summary, p-value, extraction files and sample size. |
| `reference` | LDAK, LD prefix, assembly and NCBI coordinate source. |
| `regulatory` | JEME tissue, HiC tissue/types, overlap filter and harmonization. |
| `enhancer_classification` | Disabled in the current canonical profiles. |
| `interaction_network` | Explicit `ARACNe`, `STRINGdb` or `none` backend. |
| `aracne` | Input, initial thresholds and identifier policy. |
| `ldak` | Buffers, TSS mode, result names, model and permutations. |
| `scores` | P-value column and transformation. |
| `networks` | Selected graphs, regulatory thresholds and legacy edge policies. |
| `hhotnet` | Permutation/manual mode and Python runtime. |
| `slurm` | Resource values for generated cluster scripts. |

Schema version 1 accepts the additive `interaction_network` section. Older YAML
files that omit it retain legacy ARACNe selection; new profiles should always
declare it explicitly.

For every field, implemented fallback, allowed value and affected stage, see
`vignette("analysis-yaml-reference", package = "hotnetR2")`.

## GEjunc ARACNe profile

```yaml
interaction_network:
  type: ARACNe

aracne:
  enabled: true
  file: ARACNe1600/GEjunc/network.txt
  MI_threshold: 0
  pvalue_threshold: 1.0e-6
  harmonization:
    enabled: true
    gtf_file: .cache/gencode.v39.GRCh38.genes_nodup.rds
    alias_file: .cache/alias_link_nodup.rds
    map_ensembl: true
    map_aliases: true
    ambiguous: drop
    unmapped_ensembl: drop
    unmatched_symbols: keep

networks:
  available: [network_1, network_2, network_3, network_4]
  selected: [network_1, network_2, network_3, network_4]
  thresholds:
    small_regulatory_score: 4
    broad_regulatory_score: 3
    small_plus_aracne_mi: 0.3
    broad_plus_aracne_mi: 0.65
```

Network 3 adds qualifying ARACNe edges to Network 1; Network 4 augments Network
2. Endpoints absent from the base graph are not introduced. Review the ARACNe
harmonization mapping and endpoint summaries before interpreting results.

## STRINGdb profile

```yaml
interaction_network:
  type: STRINGdb
  stringdb:
    version: "12.0"
    species: 9606
    score_threshold: 700
    cache_dir: .cache/STRINGdb
    mapping_policy: legacy_shortest_gene_per_string_id

aracne:
  enabled: false
```

Network 3 adds STRING v12 edges with `combined_score >= 700` to Network 1;
Network 4 augments Network 2. Only existing `promoter_gene` nodes are mapped.
STRING loops and duplicate gene pairs are removed; pairs already present in the
base graph are audited but do not replace regulatory edges.

The mapping policy resolves several graph gene names assigned to one STRING ID
by shortest name, with a lexical tie-break. This is deterministic compatibility,
not biological disambiguation; inspect `stringdb_mapping_conflicts.tsv`.

## Validate without running

From the project root:

```bash
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh check
```

This checks the exact installed package version requested by the launcher,
parses the YAML, validates declared inputs and verifies that output paths remain
inside the analysis folder. It does not run LDAK or HHN.

The equivalent R commands are:

```{r check}
config <- hotnetR2::read_analysis_config(
  "analysis_E092_Gastric_STRINGdb/analysis.yaml"
)
inputs <- hotnetR2::check_analysis_inputs(config, write_report = FALSE)
print(inputs, row.names = FALSE)
stopifnot(all(inputs$status == "OK"))
hotnetR2::setup_analysis(config, dry_run = TRUE)
```

## Run stages

Run one stage at a time when establishing a new profile:

```bash
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh annotations
# classification is intentionally skipped when enabled: false
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh ldak
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh scores
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh networks
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh hotnet
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh exports
```

Or run the configured sequence:

```bash
bash analysis_E092_Gastric_STRINGdb/run_analysis.sh all
```

The launcher writes a timestamped stage log. `run_analysis()` uses content
fingerprints for safe reuse, but an external tool's exit code alone is not
accepted as completion; expected artifacts are checked.

## Review before HHN

After `networks`, inspect:

```text
hHotnet/summary/network_file_summary.tsv
hHotnet/summary/promoter_harmonization_summary.tsv
hHotnet/summary/aracne_harmonization_summary.tsv       # ARACNe profile
hHotnet/summary/stringdb_summary.tsv                   # STRING profile
hHotnet/summary/stringdb_mapping_conflicts.tsv         # STRING profile
```

Also inspect annotation validation and LDAK summary manifests. A surprising
drop in mapped promoters or graph size should stop the workflow before HHN.

## Changing a profile

Changing assembly, coordinates, aliases, source selection, overlap rules,
enhancer-class policy, LDAK parameters, score column, network thresholds or
interaction backend is a scientific change. Use a new analysis name and output
directory. Do not relabel old outputs or refresh inputs inside an established
profile.

## References
