---
title: "analysis.yaml reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{analysis.yaml reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Scope and notation

This is the field-by-field reference for `analysis.yaml` schema version 1 in
`hotnetR2` 0.2.2. It documents the implemented parser, validator and runtime
consumers. The two installed examples remain the recommended starting points:

```{r examples}
example_dir <- system.file(
  "extdata", "vignette_examples", package = "hotnetR2", mustWork = TRUE
)
file.path(example_dir, c("analysis_aracne.yaml", "analysis_stringdb.yaml"))
```

The terms in the tables have precise meanings:

| Term | Meaning |
|---|---|
| Required | Omission is rejected by validation, or the field is required by a later stage. |
| Conditional | Required only when the named feature or execution mode is used. |
| Fallback | The value used by current code when the key is omitted. It is a compatibility behavior, not necessarily the recommended scientific value. |
| Example | A value in the current E092/Gastric profiles; it is not an implicit package default. |
| Inert/reserved | Accepted in YAML but currently has no computational effect. Do not use it to infer behavior. |
| Legacy fallback | Retained so older configurations still parse; new profiles should use the replacement key. |

YAML keys are case-sensitive. Enumerated values are shown using their canonical
spelling, although `interaction_network.type` is normalized to lower case.
Relative input paths resolve from `project.root`. Output paths always remain
below the directory containing `analysis.yaml`, unless `output_root` is passed
to `read_analysis_config()` for an isolated reproduction run.

The authoritative executable rules are `read_analysis_config()` and the
internal validator. This vignette intentionally identifies weak validation and
inactive fields so a future schema revision can remove or formalize them.

## Minimum structure

Schema version 1 requires these top-level sections:

```yaml
schema_version: 1
analysis: {}
project: {}
gwas: {}
reference: {}
regulatory: {}
aracne: {}
ldak: {}
scores: {}
networks: {}
hhotnet: {}
```

`interaction_network`, `enhancer_classification`, `resources`, `validation`
and `slurm` are optional sections, although new profiles should declare the
interaction backend explicitly. The empty mappings above are only structural;
the required child fields described below must also be supplied.

## Identity, paths and GWAS inputs

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `schema_version` | Required; only integer `1` | Selects the configuration schema. Other values are rejected by `read_analysis_config()`. |
| `analysis.name` | Required string | Analysis identifier used in generated job names, graph metadata and exported filenames. |
| `analysis.label` | Optional; no fallback | Descriptive metadata only; current workflow code does not consume it. |
| `analysis.random_seed` | Runtime-required integer | Passed to LDAK `--random-seed`. The validator does not yet enforce it, so omitting it can produce an invalid generated command. |
| `project.root` | Required path | Base for shared inputs and tools. A relative value is resolved from the YAML directory. It is never the output root. |
| `resources.cache_dir` | Fallback `.cache` | Base cache used by JEME and HiC readers. `read_analysis_config(cache_dir=)` overrides it. |
| `resources.profile` | Optional, inert/reserved | Written by the historical `initialize_analysis()` helper but not used by execution. |
| `gwas.dataset` | Required string | Cohort label in score summaries and provenance. |
| `gwas.summary_file` | Required path | LDAK summary-statistics input passed to `--summary`. |
| `gwas.pvalue_file` | Required path | Predictor/p-value input passed to `--pvalues`. |
| `gwas.extract_file` | Required path | Variant allow-list passed to LDAK `--extract`. |
| `gwas.sample_size` | Required positive integer | Sample size passed to LDAK `--SNP-data`. |

`check_analysis_inputs()` checks the GWAS files, LD reference components,
LDAK executable, and the selected interaction cache or network before compute.
It does not prove that column semantics or genome assembly are correct. Variant
positions in the current BEEA and MVP sources are 1-based identifiers; they
must not be treated as BED starts. See
`docs/dataset_coordinate_reference.md` in the installed source tree.

## Reference and annotation coordinates

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `reference.ldak_exe` | Required path | Executable used by generated LDAK scripts. |
| `reference.bfile_prefix` | Required path prefix | Resolves the matching `.bed`, `.bim` and `.fam` LD-reference files. |
| `reference.genome_build` | Required string | Assembly of GWAS/LD inputs. Common GRCh37 aliases are normalized. |
| `reference.annotation_genome_build` | Required string; must equal `genome_build` after normalization | Assembly of regulatory and promoter annotations. The validator prevents cross-build analysis. |
| `reference.ncbi_feature_table` | Runtime-required path | NCBI feature table used to create promoter TSS/gene-body intervals. Automatic download during analysis is deliberately disabled. |
| `reference.promoter_alias_file` | Fallback `null` | Optional two-column TSV for aliases absent from the NCBI table. `null` keeps source symbols and audits them as identity mappings. |
| `reference.promoter_alias_columns.alias` | Fallback `alias` | Alias-column name in `promoter_alias_file`. |
| `reference.promoter_alias_columns.gene_symbol` | Fallback `gene_symbol` | Approved-symbol column name in `promoter_alias_file`. Conflicting one-to-many aliases are rejected. |

`prepare_ldak_annotations()` creates 0-based, half-open `.loc` intervals. NCBI
1-based inclusive gene coordinates are converted explicitly. A positive-strand
TSS uses NCBI `START`, a negative-strand TSS uses `END`, and the one-base BED
anchor is `[TSS - 1, TSS)`. JEME enhancers and HiC restriction fragments are
already BED intervals and are not shifted again.

## JEME and HiC sources

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `regulatory.jeme.key_column` | Runtime-required | JEME metadata column used for tissue selection, normally `tiss`. |
| `regulatory.jeme.value` | Runtime-required scalar or list | Selected JEME value(s), for example `E092`. Controls network edges and the selected-JEME reference for the HiC overlap filter. |
| `regulatory.jeme.method` | Fallback `lasso` | JEME cache family loaded by `get_jeme()`. |
| `regulatory.jeme.simplified` | Fallback `true` for network loading | Selects the simplified JEME representation. Annotation preparation uses its own simplified source contract. |
| `regulatory.jeme.label` | Optional, inert/reserved | Human-readable tissue label; current stages do not consume it. |
| `regulatory.hic.enabled` | Fallback enabled unless explicitly `false` | Master HiC switch. HiC is also disabled by an absent, empty, `null` or blank tissue selection. |
| `regulatory.hic.tissue_type` | Empty disables HiC | One or more HiC tissues, for example `Gastric`. Restricts network edges; annotation preparation still uses all tissues for enabled edge types. |
| `regulatory.hic.edge_types` | Fallback `[PO, PP]`; values `PO`, `PP`, or both | Selects promoter-other and/or promoter-promoter contacts before annotation and network construction. No duplicates are allowed. |

HiC state is interpreted by the internal `analysis_hic_enabled()`,
`analysis_hic_tissues()` and `analysis_hic_edge_types()` helpers. The public
data-reader meaning of `get_hic(NULL)` remains “all tissues”; the empty-value
disable rule applies only to analysis configuration.

### Optional HiC--JEME overlap filter

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `regulatory.hic.jeme_overlap.enabled` | Required if the subsection is present | When `true`, removes HiC PO enhancers that overlap the selected JEME enhancer union. PP contacts are unchanged. |
| `regulatory.hic.jeme_overlap.hic_flank_bp` | Fallback `1000`; nonnegative integer | Expands each HiC enhancer by this many bases at both ends for overlap testing. |
| `regulatory.hic.jeme_overlap.jeme_flank_bp` | Fallback `1000`; nonnegative integer | Expands each selected JEME enhancer by this many bases at both ends. |

The filter uses 0-based, half-open intersection; touching boundaries do not
overlap. It runs for both the annotation and network source loads and writes
separate audit tables. It requires `bedtools` when enabled.

### Promoter identifier harmonization

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `regulatory.promoter_harmonization.enabled` | Fallback `false` | Enables canonical promoter mapping before annotation and network construction. Current profiles set `true`. |
| `.jeme` | Fallback `true` when harmonization is enabled | Applies the JEME mapping. |
| `.hic` | Fallback `true` when harmonization is enabled | Applies the HiC symbol/alias mapping. Has no HiC effect if HiC is disabled. |
| `.jeme_strategy` | Fallback and only value `all_tissues_ensg_map` | Builds one ENSG-to-symbol map across all JEME tissues, then applies it consistently to all-tissue annotations and selected-tissue edges. |
| `.hic_strategy` | Fallback and only value `gene_symbol_alias` | Harmonizes HiC PO/PP promoter symbols through the alias reference. |
| `.gtf_file` | Conditional path | Required when harmonization is enabled; RDS must provide the expected GENCODE gene mapping. It is used for identifiers, not GRCh38 coordinates. |
| `.hic_alias_file` | Conditional path | Required when harmonization and HiC are enabled; RDS contains the curated alias mapping. |

## MHC exclusion

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `regulatory.mhc_exclusion.enabled` | Fallback `false` | If enabled, removes every node type overlapping the fixed conservative MHC interval and all incident regulatory edges. |
| `.genome_build` | Fixed `hg19`/GRCh37 | Optional explicit assertion; any other assembly is rejected. |
| `.chr` | Fixed `6` | Optional explicit assertion. |
| `.start` | Fixed `25000000` | 1-based inclusive source-region start, converted internally for BED overlap. |
| `.end` | Fixed `34000000` | 1-based inclusive source-region end. |

The region cannot be narrowed through YAML. Legacy `enhancers` or `promoters`
switches are rejected because exclusion is deliberately all-node-type.

## Enhancer classification

The current E092/Gastric profiles set `enhancer_classification.enabled: false`:
they use corrected TSS annotations with no enhancer-class exclusion. The keys
below preserve a historical alternative and should not be copied into the
canonical profile as if they were defaults.

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `enhancer_classification.enabled` | Fallback `false` | Enables `classify_enhancers()` and class-dependent retention. |
| `.source_analysis` | Conditional string, currently inert/reserved | Required by validation when enabled, but not read by classification or execution. |
| `.enhancer_flank_bp` | Conditional nonnegative integer | Expands enhancer BED intervals during enhancer-versus-promoter classification. |
| `.promoter_window` | Legacy fallback `gene_body`; values `gene_body`, `strand_aware_tss` | Labels the classification profile and controls the fallback source-retention policy. It should agree with `ldak.promoter_interval_mode`. |
| `.promoter_flank_bp` | Conditional nonnegative integer | Expands the chosen promoter intervals for classification. |
| `.gene_node_flank_bp` | Legacy fallback for `promoter_flank_bp` | Older spelling accepted by `read_analysis_config()`; do not use in new YAML. |
| `.class1_label`, `.class2_label`, `.class3_label` | Conditional, unique safe identifiers | Names no-overlap, fully-contained and partial-overlap classes, respectively. |
| `.drop_non_class1_edge_sources` | Optional list from `JEME`, `HiC_PO` | Explicitly drops class2/3 edges from listed sources. Empty `[]` retains all classes. If omitted, TSS mode drops only `HiC_PO`; gene-body mode drops both. |
| `.retain_non_enhancer_edges` | Inert/reserved | Present in older templates but not consumed by current code. |

Classification uses `bedtools`; it writes complete class and overlap audits plus
the retained enhancer `.loc` input. The run sequence must skip the
`classification` stage when classification is disabled.

## Interaction backend

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `interaction_network.type` | Recommended explicit value: `ARACNe`, `STRINGdb`, or `none` | Selects the backend used to augment Networks 1/2 into Networks 3/4. |

If this key is omitted, legacy behavior selects ARACNe when
`aracne.enabled` is omitted or true, and `none` when it is false. This fallback
is for old files; new profiles should always state the backend. With `none`,
`networks.available` cannot include Network 3 or 4.

### STRINGdb settings

| Key | Fallback or allowed values | Effect and related code |
|---|---|---|
| `interaction_network.stringdb.version` | `"12.0"` | Pins the STRING release and expected cache filenames. Quote it so YAML preserves the version string. |
| `.species` | `9606` | NCBI taxonomy ID; current examples use human. |
| `.score_threshold` | `700`; numeric 0--1000 | Inclusive minimum `combined_score`. Networks 3/4 receive qualifying STRING pairs between existing promoter-gene nodes. |
| `.cache_dir` | `.cache/STRINGdb` | STRING-specific cache root, resolved from `project.root`; it is independent of `resources.cache_dir`. |
| `.mapping_policy` | Only `legacy_shortest_gene_per_string_id` | Resolves multiple graph genes mapped to one STRING ID by shortest name then lexical tie-break. Conflicts are audited. |

`download_stringdb_data()` is an explicit setup operation. Analysis execution
never downloads missing STRING files. Loops, duplicate pairs and pairs already
represented by a regulatory edge are handled and reported by the network
builder; STRING does not introduce enhancer or new gene nodes.

### ARACNe settings

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `aracne.enabled` | Legacy fallback `true` | Must be true for the ARACNe backend and false for STRINGdb or `none`. |
| `aracne.file` | Conditional path | Required when ARACNe is enabled. |
| `aracne.MI_threshold` | Fallback `0` | Initial inclusive MI filter in `read_ARACNe()`. Separate Network 3/4 MI thresholds are applied later. |
| `aracne.pvalue_threshold` | Fallback `1e-6` | Initial inclusive maximum p-value filter. |
| `aracne.keep_genes` | Fallback `null` | Optional vector used by `read_ARACNe()` to retain edges involving named genes in addition to threshold-passing edges. |
| `aracne.harmonization.enabled` | Fallback `true` | Enables endpoint identifier harmonization. |
| `.gtf_file` | Fallback `.cache/gencode.v39.GRCh38.genes_nodup.rds` | ENSG and canonical-symbol mapping input. |
| `.alias_file` | Fallback `.cache/alias_link_nodup.rds` | Curated alias mapping input. |
| `.map_ensembl` | Fallback `true` | Enables ENSG mapping. |
| `.map_aliases` | Fallback `true` | Enables alias mapping. |
| `.ambiguous` | Only `drop` | Drops identifiers with multiple canonical candidates. |
| `.unmapped_ensembl` | Only `drop` | Drops unresolved ENSG endpoints. |
| `.unmatched_symbols` | Only `keep` | Retains unmatched non-ENSG symbols unchanged. |

`harmonize_aracne_endpoints()` writes mapping, endpoint and edge summaries.
Disabling harmonization is permitted, but the path fields are still normalized
by the current settings helper and the resulting identifiers must already agree
with graph gene names.

## LDAK annotations and jobs

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `ldak.enhancer_flank_bp` | Required in new profiles; nonnegative integer | LDAK region flank for enhancer annotations. |
| `ldak.promoter_flank_bp` | Required in new profiles; nonnegative integer | LDAK region flank for promoter annotations. |
| `ldak.flank_bp` | Legacy common fallback | Used only when a node-specific flank is absent. Do not use in new profiles. |
| `ldak.promoter_interval_mode` | Fallback `gene_body`; values `gene_body`, `strand_aware_tss` | Chooses NCBI gene-body intervals or one-base strand-aware TSS anchors. Current profiles use `strand_aware_tss`. |
| `ldak.result_dirs.enhancer` | Fallback `enhancer` | Safe subdirectory below `LDAK/results`; change it when results are scientifically incompatible. |
| `ldak.result_dirs.promoter` | Fallback `promoter` | Equivalent promoter-result subdirectory. |
| `ldak.annotation_tissues.jeme` | Fallback `ALL`; currently only `ALL` | Source universe for LDAK annotations. Tissue-restricted annotation preparation is not implemented. |
| `ldak.annotation_tissues.hic` | Fallback `ALL`; currently only `ALL` when HiC enabled | HiC annotation universe. Selected tissues still restrict network edges. |
| `ldak.annotation_tissues.jeme_method` | Fallback `lasso` | JEME method used for the all-tissue annotation universe. |
| `ldak.power` | Runtime-required numeric | Passed to LDAK `--power`; current examples use `-0.25`. |
| `ldak.pruning` | Runtime-required numeric | Passed to `--gene-prune`; current examples use `0.95`. |
| `ldak.permutations` | Required nonnegative integer | Passed to `--gene-permutations`; current examples use `10`. |
| `ldak.threads` | Required positive integer | Default thread count in generated LDAK scripts; current examples use `8`. |
| `ldak.ignore_weights` | Required boolean | Emits LDAK `--ignore-weights YES/NO`. |
| `ldak.allow_ambiguous` | Required boolean | Emits LDAK `--allow-ambiguous YES/NO`. |

`prepare_ldak_annotations()`, `generate_ldak_jobs()`, `run_ldak()` and
`validate_ldak_results()` are the main consumers. LDAK result reuse is guarded
by input/configuration fingerprints. A symbolic-link result directory is never
used as a write target.

## Score conversion

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `scores.pvalue_column` | Required string | Column read from LDAK output, normally `LRT_P_Perm`. |
| `scores.transform` | Required; only `neg_log10` | Node score is `-log10(p)`. |
| `scores.infinite_score_cap` | Fallback `null`, or positive number | Replaces positive infinity from `p = 0`. With `null`, the cap is one above the largest finite score, or 1 when none is finite. |

These values are consumed by `summarize_analysis_ldak_results()` and
`build_hhotnet_networks()`. P-values outside `[0, 1]` are rejected.

## Network construction

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `networks.available` | Legacy fallback to `selected` | Complete buildable set; unique subset of `network_1` through `network_4`. |
| `networks.selected` | Required nonempty list | Default set used when a function receives no explicit `networks` argument. Must be a subset of `available`. |
| `networks.regulatory_edge_filter` | Required; `current` or `legacy_filter_network_ldak` | `current` removes incomplete edges and exact duplicates before score/node filtering. The legacy mode preserves historical row behavior. |
| `networks.node_order` | Required; `score_descending` or `legacy_input` | Controls numeric node IDs written for HHN. Gene membership and scores are unchanged, but file order and IDs differ. |
| `networks.preserve_self_loops` | Fallback `true` | Keeps or removes regulatory and interaction self-loops before graph creation. |
| `networks.deduplicate_jeme_edges_before_harmonization` | Required boolean | If true, collapses duplicate enhancer--promoter JEME rows before harmonization. |
| `networks.split_multi_promoter_edges` | Required boolean | Splits semicolon-delimited promoter targets into separate edges. |
| `networks.preserve_duplicate_edges_during_aracne` | Conditional boolean | ARACNe-only compatibility policy for duplicate augmented edges. |
| `networks.thresholds.small_regulatory_score` | Required nonnegative number | Network 1 retains regulatory components whose maximum node score meets this cutoff; Network 3 starts from that graph. Current example: `4`. |
| `.broad_regulatory_score` | Required nonnegative number | Equivalent cutoff for Networks 2/4. Current example: `3`. |
| `.small_plus_aracne_mi` | ARACNe-only required nonnegative number | Minimum MI for ARACNe edges added to Network 1 to form Network 3. Current example: `0.3`. |
| `.broad_plus_aracne_mi` | ARACNe-only required nonnegative number | Minimum MI for ARACNe edges added to Network 2 to form Network 4. Current example: `0.65`. |

For STRINGdb, Networks 3/4 instead use the single STRING score threshold. For
`none`, advertise only Networks 1/2. `available_networks()` and
`resolve_networks()` expose the resolved sets; `build_hhotnet_networks()` writes
the graphs, index tables, scores and audit summaries.

## Hierarchical HotNet

| Key | Requirement or fallback | Effect and related code |
|---|---|---|
| `hhotnet.analysis_mode` | Fallback `permutation`; `permutation` or `manual_delta` | Standard permutation inference, or observed hierarchy with user-selected delta cuts. `manual_delta` requires `local_python`. |
| `hhotnet.execution_mode` | Fallback `apptainer`; `apptainer` or `local_python` | Selects the HHN runtime used by generated scripts. |
| `hhotnet.local_dir` | Required for `local_python` | Hierarchical HotNet checkout containing `src/`. |
| `hhotnet.python_executable` | Fallback `python3` | Python command for local execution. |
| `hhotnet.compat_dir` | Fallback `hHotnet/compat` below the analysis output | Python compatibility shim directory. A relative explicit path resolves from `project.root`. |
| `hhotnet.apptainer_image` | Runtime-required for `apptainer` | Container image. It may remain as unused metadata in a local-Python profile. |
| `hhotnet.score_name` | Runtime-required; related readers fall back to `nodes_all` | Basename of the score table and part of HHN result paths. New profiles should set it explicitly to `nodes_all`. |
| `hhotnet.num_permutations` | Required nonnegative integer | Number of score/hierarchy permutations in standard mode. Manual-delta mode stops before permutations. |
| `hhotnet.permutation_seed` | Fallback `0`; nonnegative integer | Base seed for score permutations. |
| `hhotnet.lower_size_bound` | Required positive integer | Minimum cluster size supplied to permutation hierarchy processing. |
| `hhotnet.cpus` | Required positive integer | Parallelism for HHN jobs and generated Slurm CPU request. |
| `hhotnet.delta_values` | Fallback `[0.05, 0.1, 0.2, 0.5]` in `run_analysis()` | Manual-delta cuts of the observed hierarchy. They are exploratory and have no permutation significance. |

The main consumers are `generate_hhotnet_job()`, `run_hhotnet()`,
`extract_hhotnet_delta_clusters()`, `summarize_hhotnet_results()` and
`export_hhotnet_graphs()`. Environment variables beginning `HHOTNET_` can
override selected generated-script values; they are included in reuse
fingerprints.

## Slurm and optional reference comparisons

| Key | Fallback | Effect and related code |
|---|---|---|
| `slurm.time` | LDAK `12:00:00`; HHN `72:00:00` | Generated `#SBATCH --time`. The current examples explicitly use `12:00:00` for both. |
| `slurm.memory` | LDAK `16G` | LDAK generated memory request. The HHN generator does not currently emit a memory directive. |
| `slurm.partition` | `general-compute` | Generated partition for both job types. |
| `slurm.qos` | `general-compute` | Generated QoS for both job types. |
| `validation.reference_ldak_dir` | `null` | Optional read-only historical result root used by LDAK comparison helpers. It does not validate routine runs automatically. |
| `validation.reference_hhotnet_dir` | `null` | Optional read-only historical `hHotnet` root used by `compare_hhotnet_networks()`. |

Slurm keys affect generated scripts, not statistical results, except that CPU
settings can affect runtime. Submission remains a user/cluster operation;
`hotnetR2` does not call `sbatch` automatically.

## Functions related to configuration

| Function | Role |
|---|---|
| `read_analysis_config()` | Reads YAML, applies limited compatibility normalization, validates it, and resolves root/cache metadata. |
| `check_analysis_inputs()` | Reports whether required files for the chosen backend exist. |
| `analysis_paths()` | Returns guarded analysis-local output directories. |
| `setup_analysis()` | Creates the output structure and records configuration/provenance. |
| `write_effective_config()` | Writes resolved roots, paths and network sets to `metadata/effective_analysis.yaml`. It does not currently materialize every fallback value. |
| `available_networks()` / `resolve_networks()` | Resolve the advertised and default network sets. |
| `prepare_ldak_annotations()` | Consumes reference, regulatory, harmonization, MHC and LDAK interval settings. |
| `classify_enhancers()` | Consumes the optional enhancer-classification policy. |
| `generate_ldak_jobs()` / `run_ldak()` | Consume GWAS, LD, LDAK, analysis seed and Slurm settings. |
| `summarize_analysis_ldak_results()` | Converts LDAK outputs into enhancer/promoter score tables and creates the all-tissue JEME long enhancer report. |
| `create_enhancer_ldak_long_summary()` | Recreates `enhancer_ldak_long.tsv.gz` from an existing enhancer summary and JEME cache without rerunning LDAK. |
| `build_hhotnet_networks()` | Consumes source, score, backend and network policies. |
| `generate_hhotnet_job()` / `run_hhotnet()` | Consume runtime, permutation and Slurm settings. |
| `extract_hhotnet_delta_clusters()` | Applies manual delta values to an observed hierarchy. |
| `run_analysis()` | Runs selected stages in order; it does not automatically suppress `classification` when disabled. |

The low-level validator is intentionally internal. Calling
`read_analysis_config()` is the supported validation interface.

## Change impact: what must be regenerated

| Changed setting | Earliest affected stage | Regenerate downstream |
|---|---|---|
| Assembly, NCBI table, JEME/HiC selection, harmonization, MHC, overlap filter, annotation tissue universe | `annotations` | Classification if used, LDAK, scores, networks, HHN, exports |
| Enhancer-class flanks/window/retention | `classification` | Enhancer LDAK, scores, networks, HHN, exports |
| LDAK flank, model, pruning, permutations, ambiguity policy, GWAS or LD input | `ldak` | Scores, networks, HHN, exports |
| Score p-value column or infinity cap | `scores` | Networks, HHN, exports |
| Regulatory thresholds, edge policies, node order, ARACNe/STRING backend or thresholds | `networks` | HHN and exports |
| HHN runtime, permutation count/seed, lower bound or analysis mode | `hotnet` | HHN results and exports |
| Manual `delta_values` only | `exports` after a valid observed hierarchy | Manual-delta tables and graph exports |
| Slurm scheduling fields only | job generation | Regenerate scripts; scientific inputs are unchanged |

For any scientific change, use a new analysis directory/name and new result
subdirectories. Do not reinterpret old artifacts under an edited YAML file.

## Compatibility and future schema cleanup

The following are candidates for a later schema revision:

- remove implicit ARACNe selection and require `interaction_network.type`;
- remove `ldak.flank_bp` and `enhancer_classification.gene_node_flank_bp`;
- either implement or remove `resources.profile`, `analysis.label`,
  `regulatory.jeme.label`, `enhancer_classification.source_analysis` and
  `enhancer_classification.retain_non_enhancer_edges`;
- enforce runtime-required `analysis.random_seed`, `reference.ncbi_feature_table`,
  `ldak.power`, `ldak.pruning`, `hhotnet.score_name` and the mode-specific
  `hhotnet.apptainer_image` during initial validation;
- materialize all effective fallback values, not only paths and network sets,
  in the provenance YAML;
- introduce a new `schema_version` when a removal or changed default would make
  the same YAML produce a different analysis.

Until such a versioned change is made, do not silently alter the fallbacks in
this document. Add a validator regression test, update both installed example
YAML files where relevant, record the decision in `PROJECT_SPEC.md` and
`NEWS.md`, and describe whether existing analyses must be regenerated.
