---
title: "Current corrected TSS workflow"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Current corrected TSS workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Canonical scope

This vignette describes the recommended E092/Gastric profile. It borrows the
four-network organization and HHN execution patterns from v6/v7, but it does not
claim byte identity with their frozen outputs. Corrected coordinates, no
enhancer-class exclusion and current curated aliases are intentional changes.

## Regulatory sources

- JEME E092 supplies fetal-stomach enhancer--target edges [@Cao2017JEME].
- Jung Gastric HiC supplies promoter--other and promoter--promoter contacts
  [@Jung2019HiC].
- All JEME tissues define the ENSG harmonization universe; E092 defines the
  selected network edges.
- HiC promoters are harmonized by gene symbol and curated aliases.

The optional HiC--JEME filter compares selected-tissue JEME enhancers to HiC
promoter--other fragments with declared flanks. Positive-width overlap removes
the HiC PO contact. JEME and HiC PP edges are unchanged.

## Correct coordinate conversion

All active `.loc` and bedtools intervals are 0-based, half-open.

For an NCBI GRCh37 record with 1-based inclusive `START`, `END` and strand:

- positive-strand TSS is `START`;
- negative-strand TSS is `END`;
- the one-base TSS interval is `[TSS-1,TSS)`;
- the gene-body interval, when a historical profile explicitly requests it, is
  `[START-1,END)`.

JEME enhancer bounds and Jung HiC other-end fragments are already BED-style.
Their starts pass through without another subtraction. Touching half-open
boundaries are adjacent, not overlapping.

The `.loc` output is headerless:

```text
name  CHR  START  END
```

It contains unexpanded coordinates. LDAK receives enhancer and promoter buffers
separately through `--gene-buffer`.

## No enhancer-class exclusion

The canonical YAML declares:

```yaml
enhancer_classification:
  enabled: false
```

Thus the classification stage is skipped by `run_analysis.sh all`, and no edge
is removed because of an enhancer class. This is distinct from older gene-body
or TSS profiles that classified expanded enhancers against promoter windows and
dropped configured sources. Re-enabling classification is a separate scientific
profile requiring regenerated annotations, LDAK scores, networks and expected
results.

The selected-JEME versus HiC-PO overlap filter is independent of enhancer-class
exclusion and may remain enabled.

## Promoter harmonization

```yaml
regulatory:
  promoter_harmonization:
    enabled: true
    jeme: true
    hic: true
    jeme_strategy: all_tissues_ensg_map
    hic_strategy: gene_symbol_alias
    gtf_file: .cache/gencode.v39.GRCh38.genes_nodup.rds
    hic_alias_file: .cache/alias_link_nodup.rds
```

JEME target ENSG IDs are resolved across all tissues before the selected E092
edges are constructed. HiC endpoints use symbol/alias mapping. The current
curated references mark the mapping direction; unmarked frozen caches must not
be silently reinterpreted or refreshed.

Promoter coordinates are subsequently rematched to the NCBI GRCh37 feature
table. Neither the raw JEME TSS nor GENCODE GRCh38 positions replace those
current LDAK anchors.

## LDAK stages and scores

For each enabled node type, hotnetR2:

1. writes and validates the unexpanded `.loc` file;
2. runs `--cut-genes` with the configured buffer and p-value file;
3. runs `--calc-genes-reml` with the configured summary statistics;
4. runs `--join-genes-reml`;
5. checks expected artifacts before accepting external exit status;
6. standardizes enhancer and promoter summary tables;
7. expands enhancer scores over all unique JEME target/tissue associations in
   `enhancer_ldak_long.tsv.gz`.

The current profile uses LDAK-GBAT [@Berrandou2023LDAK] and scores nodes as
`-log10(LRT_P_Perm)`. With a configured infinity cap, positive infinity becomes
the maximum finite score plus one. P-values outside `[0,1]` are invalid.

Standard summaries use `Gene_Name` as their sole node identifier; the former
duplicate `gene` column is not written. The standard enhancer table adds
`original_promoter`, collapsing sorted distinct raw JEME promoter labels with
semicolons for each enhancer. The long enhancer table keeps one
`original_promoter` per association together with raw JEME `promoterFull` and
`ENSG`, harmonized `promoter`, `tissue` (for example `E092`) and `tissue_name`
(for example `Fetal Stomach`). It retains HiC-only enhancers with missing JEME
fields. For a completed analysis, upgrade the standard enhancer summary and
recreate the long report without running LDAK:

```{r enhancer-long}
config <- hotnetR2::read_analysis_config("analysis_name/analysis.yaml")
hotnetR2::create_enhancer_ldak_long_summary(config)
```

## Regulatory networks

JEME, HiC PO and HiC PP edges are combined after source-specific filtering and
harmonization. Network thresholds retain whole undirected connected components:

- Network 1: component maximum score at least 4;
- Network 2: component maximum score at least 3.

Weak connector nodes can therefore remain. These thresholds are not per-node
filters. The current profiles retain the established input ordering and
self-loop policy; do not simplify edges unless the profile explicitly changes.

Networks 3/4 add the selected interaction backend only between endpoints already
present in Network 1/2. ARACNe and STRINGdb policies are described in
`vignette("analysis-profiles-and-running", package = "hotnetR2")`.

## Audits to review

Before HHN, inspect at least:

- annotation and interval-validation summaries;
- JEME ENSG and HiC alias mapping tables;
- unmatched enhancer/promoter scores;
- selected-JEME/HiC overlap summaries;
- MHC exclusion counts when enabled;
- ARACNe or STRING mapping/conflict/edge summaries;
- network node and edge counts.

Do not interpret a clean software exit as proof of correct resource identity or
scientific equivalence to an older run.

## References
