---
title: "Hierarchical HotNet execution modes"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Hierarchical HotNet execution modes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Inputs to Hierarchical HotNet

Hierarchical HotNet (HHN) combines graph topology with vertex scores to identify
a hierarchy of altered subnetworks [@Reyna2018HHN]. For every selected network,
hotnetR2 writes:

- a contiguous 1-based node index;
- a two-column integer edge list;
- a named node-score table;
- full node and edge audit tables.

HHN may restrict hierarchy construction to a largest strongly connected
component. Therefore, not every input graph node is guaranteed to appear in the
reported hierarchy or clusters.

## v6-style permutation mode

Permutation mode is the standard inference workflow:

```yaml
hhotnet:
  analysis_mode: permutation
  execution_mode: local_python
  score_name: nodes_all
  num_permutations: 100
  permutation_seed: 0
  lower_size_bound: 10
```

The observed score vector has index 0. The local execution profile uses base
seed plus the permutation index, so base 0 produces seeds 1 through 100.
Observed and null hierarchies are used to select and assess cluster results.

LDAK permutations and HHN permutations are different procedures. Setting HHN
`num_permutations` to zero does not disable the LDAK permutations configured in
the `ldak` section.

## v7-style manual-delta mode

Manual-delta mode is an exploratory observed-hierarchy workflow:

```yaml
hhotnet:
  analysis_mode: manual_delta
  delta_values: [0.05, 0.1, 0.2, 0.5]
  execution_mode: local_python
  score_name: nodes_all
  num_permutations: 0
  permutation_seed: 0
  lower_size_bound: 10
```

It is currently supported only with `local_python`. The `hotnet` stage stops
after similarity matrices are built. The `exports` stage constructs or reuses
the observed hierarchy and cuts it at the requested delta values using HHN's
native semantics.

```{r manual-cuts}
config <- hotnetR2::read_analysis_config("analysis.yaml")
hotnetR2::extract_hhotnet_delta_clusters(
  config,
  deltas = c(0.05, 0.1, 0.2, 0.5),
  networks = NULL,
  min_cluster_size = 2L
)
```

Manual cuts have `significance: not_evaluated`. They are not permutation
p-values, do not select an optimal delta, and should not be presented as
statistically significant subnetworks.

## Output differences

| Property | Permutation mode | Manual-delta mode |
|---|---|---|
| Null hierarchies | Yes | No |
| Significance inference | Standard configured HHN workflow | Not evaluated |
| Cluster files | Selected standard result | One set per delta |
| Singleton handling | Standard exporter rules | Raw tables retain singletons |
| CX2 membership | May include surrounding connected component | Qualifying cluster members and within-cluster edges only |
| Output location | Standard result/cytoscape paths | `manual_delta` subdirectories |

Do not refactor these exporters into one membership rule. Their scopes are
intentionally different.

## Completion checks

```{r completion}
config <- hotnetR2::read_analysis_config("analysis.yaml")
hotnetR2::validate_hhotnet_inputs(config)
hotnetR2::validate_hhotnet_completion(config)
```

In manual mode, `validate_hhotnet_completion()` establishes similarity-matrix
presence, not successful extraction of every delta. Verify the manual manifest
and each referenced cluster, node, edge and CX2 file separately.

The manual manifest describes the latest invocation. Older delta files may
remain, so the directory alone is not a run inventory. Cluster numbers are
output-local; compare exact membership, not cluster number, across analyses.

## Reuse and reproducibility

hotnetR2 fingerprints content used by package-owned stages. Changed or unstamped
result directories can be moved to `superseded-*` rather than overwritten.
Runtime identity still matters: record the HHN commit, Python packages, compiled
extension, platform and seeds. Existing similarity matrices are not, by
themselves, proof that they match a changed network.

## References
