---
title: "Input preprocessing for hotnetR2"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Input preprocessing for hotnetR2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Scope and status

`hotnetR2` consumes processed GWAS, LD-reference and optional ARACNe files. It
does not prescribe one raw-cohort QC pipeline. Allele QC, sample-size rules,
liftover, duplicate handling and imputation filters are cohort-specific and
must be recorded upstream.

> **Current limitation:** complete raw-to-processed recipes for the BEEA GWAS
> and GEjunc ARACNe data are not yet certified. The schemas and checks below are
> current input contracts; example transformations are templates, not claims
> that they reproduce the frozen datasets.

## Assembly and coordinate contract

The current E092/Gastric profiles use GRCh37/hg19.

- GWAS and PLINK BIM positions are 1-based single-base variant identifiers.
- A predictor is written as `chromosome:position`; do not subtract one.
- LDAK `.loc` and bedtools intervals are 0-based, half-open.
- Convert a 1-based inclusive interval `[start,end]` to `[start-1,end)` once.
- JEME enhancer bounds and Jung HiC other-end fragments are already BED-style;
  do not decrement their starts again.
- GENCODE v39 GRCh38 coordinates are used only for identifier harmonization in
  these profiles.

See `docs/dataset_coordinate_reference.md` in the source package for the
dataset-specific evidence and cache hashes.

## GWAS files

The YAML declares three files:

```yaml
gwas:
  dataset: MY_COHORT
  summary_file: GWAS/MY_COHORT/ldak_MY_COHORT.tsv
  pvalue_file: GWAS/MY_COHORT/ldak_MY_COHORT_pvalue.tsv
  extract_file: GWAS/MY_COHORT/ldak_MY_COHORT_extract.txt
  sample_size: 50000
```

### Summary statistics

The current analyses use an LDAK summary file with this header:

```text
Predictor  A1  A2  Z  n
10:100000625  A  G  -1.53691  49266
```

`Predictor` must match the LD-reference BIM identifier exactly. `A1` and `A2`
are alleles, `Z` is the signed test statistic, and `n` is the effective sample
size used for that record. Other LDAK-supported summary formats require an
explicitly tested profile rather than a silent column substitution.

### P-value file

The current LDAK `--cut-genes` input is:

```text
Predictor  P
10:100000625  0.1237
```

P-values must be numeric and within `[0,1]`. Each predictor must use the same ID
and assembly as the summary file and reference BIM.

### Extraction list

The extraction file is headerless, one predictor per line:

```text
10:100000625
10:100000645
```

It should be the allele-compatible intersection between the processed GWAS and
the chosen LD reference. Coordinate matching and allele orientation are separate
checks: compare direct, swapped, complemented and complement-swapped alleles,
and define an explicit policy for ambiguous A/T and C/G variants.

### Cohort-specific preparation template

Replace the placeholder column names and document every filter. This example is
deliberately not a BEEA reproduction script.

```{r gwas-template}
library(dplyr)
library(readr)

raw <- read_tsv("GWAS/MY_COHORT/raw_summary.tsv.gz")

prepared <- raw |>
  transmute(
    chromosome = CHROMOSOME_COLUMN,
    position = POSITION_GRCH37_COLUMN, # 1-based
    Predictor = paste0(chromosome, ":", position),
    A1 = EFFECT_ALLELE_COLUMN,
    A2 = OTHER_ALLELE_COLUMN,
    Z = EFFECT_COLUMN / STANDARD_ERROR_COLUMN,
    n = EFFECTIVE_SAMPLE_SIZE_COLUMN,
    P = PVALUE_COLUMN
  ) |>
  filter(
    chromosome %in% c(1:22, "X"),
    is.finite(Z), is.finite(n), n > 0,
    is.finite(P), P >= 0, P <= 1
  )

# Next: exact BIM-ID and allele-orientation matching. Do not use position alone.
stop("PLACEHOLDER: implement and audit cohort-specific BIM allele matching")
```

Save final tables without row names or quoting:

```{r write-gwas}
write_tsv(select(prepared, Predictor, A1, A2, Z, n),
          "GWAS/MY_COHORT/ldak_MY_COHORT.tsv")
write_tsv(select(prepared, Predictor, P),
          "GWAS/MY_COHORT/ldak_MY_COHORT_pvalue.tsv")
write_lines(prepared$Predictor,
            "GWAS/MY_COHORT/ldak_MY_COHORT_extract.txt")
```

## LD reference

Supply one PLINK binary prefix:

```text
LD_ref/reference.bed
LD_ref/reference.bim
LD_ref/reference.fam
```

and configure:

```yaml
reference:
  bfile_prefix: LD_ref/reference
```

The `.bim` file has six fields: chromosome, variant ID, genetic distance,
1-based base-pair position, allele 1 and allele 2. The reference ancestry and
assembly should be appropriate for the GWAS. Validate that every extraction ID
occurs in the BIM and that the three PLINK files share the same prefix.

```{r ld-check}
bim <- read_table2(
  "LD_ref/reference.bim", col_names = FALSE,
  col_types = cols(.default = col_character())
)
names(bim) <- c("CHR", "Predictor", "CM", "BP", "A1", "A2")
extract <- read_lines("GWAS/MY_COHORT/ldak_MY_COHORT_extract.txt")
stopifnot(!anyDuplicated(bim$Predictor))
stopifnot(all(extract %in% bim$Predictor))
```

## ARACNe network

ARACNe reconstructs regulatory associations from expression data
[@Margolin2006ARACNe; @Lachmann2016ARACNeAP]. `hotnetR2` expects a tab-separated
file with a header and four columns:

```text
Regulator  Target  MI  pvalue
CDX1  ENSG00000277351  0.2515694401671866  0
```

- `Regulator` and `Target` may contain gene symbols or Ensembl gene IDs.
- `MI` and `pvalue` must be numeric.
- Profile thresholds are applied before endpoint harmonization.
- Current GEjunc settings initially use `MI >= 0` and `pvalue <= 1e-6`; Networks
  3/4 then use their configured MI thresholds.

The package contains helpers for exporting tissue TPM data and producing an
ARACNe job, but the exact GTEx release, tissue, gene types, transformation,
transcription-factor list, adaptive-partitioning version, bootstrap count,
randomness and consolidation environment must be frozen for a reproducible
network.

```{r aracne-template}
# Illustrative only; confirm the installed function help and site environment.
hotnetR2::download_gtex_tpm(
  tissue = "Esophagus_Gastroesophageal_Junction"
)
hotnetR2::download_gtex_gencode()
hotnetR2::export_gtex_aracne(
  tissue = "Esophagus_Gastroesophageal_Junction",
  output_dir = "ARACNe1600/GEjunc_work"
)
hotnetR2::create_aracne_slurm(
  tissue = "Esophagus_Gastroesophageal_Junction",
  output_dir = "ARACNe1600/GEjunc_work",
  aracne_jar = "/path/to/aracne.jar"
)
```

After external execution, normalize the consolidated network to the four-column
schema and preserve the raw network, commands, logs, software identity and
checksums. Do not substitute an all-gene network for a coding/lincRNA/miRNA
profile without declaring a new analysis.

## STRINGdb alternative

STRINGdb does not require ARACNe preprocessing. Use the explicit setup described
in `vignette("resources-and-setup", package = "hotnetR2")`. The aliases,
protein-info and links files must all use the same release and species.

## Preflight

Before analysis, load the YAML and require every declared input to be present:

```{r preflight}
config <- hotnetR2::read_analysis_config("analysis_name/analysis.yaml")
check <- hotnetR2::check_analysis_inputs(config, write_report = FALSE)
print(check, row.names = FALSE)
stopifnot(all(check$status == "OK"))
```

Presence is not proof of provenance. Keep checksums and upstream decisions in an
immutable run manifest.

## References
