---
title: "Reproduction and migration boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproduction and migration boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Current versus frozen behavior

The recommended E092/Gastric profiles use corrected 0-based half-open
coordinates, strand-aware NCBI TSS anchors, no enhancer-class exclusion and
current curated harmonization references. These are intentional scientific
changes from several historical result folders.

Frozen v6/v7 outputs remain valuable for their orchestration patterns:

- v6 demonstrates four networks plus standard HHN permutations;
- v7 demonstrates the same upstream network organization followed by an
  observed hierarchy and manual delta cuts.

They are not byte-level reference outputs for corrected coordinates or refreshed
aliases. Never overwrite the frozen folders when testing current behavior.

## What to freeze

A reproducible run records at least:

- `analysis.yaml` and its fully resolved effective copy;
- source package version and source/tarball SHA-256;
- hashes of GWAS, LD, JEME, HiC, NCBI, GENCODE/alias and interaction inputs;
- LDAK, bedtools, Python and Hierarchical HotNet identities;
- seeds, threads and execution mode;
- analysis-local logs, manifests and audit tables.

`check_analysis_inputs()` verifies declared presence, not complete provenance.
`capture_resources()` and `provision_resources()` can verify a declared local
resource bundle using SHA-256.

```{r provision}
hotnetR2::provision_resources(
  "resources.tsv", root = "/path/to/new-project",
  source_root = "/path/to/frozen-bundle"
)
```

## Comparison scope

Declare which artifacts are expected to be identical. Deterministic package
outputs can be compared byte-for-byte; gzip content can be compared after
decompression. Absolute paths, timestamps, environment reports and external
stochastic results require separately justified comparisons.

```{r compare}
hotnetR2::compare_artifacts(
  reference = "/path/to/frozen-analysis",
  candidate = "/path/to/new-analysis",
  files = readLines("artifact_inventory.txt")
)
```

Matching selected files or counts is not whole-run equivalence. In particular,
fresh LDAK permutation-adjusted p-values have shown external variability under
some historical settings even when observed statistics agree.

## Safe migration procedure

1. Restore immutable inputs into a new project.
2. Copy a package-local current example; do not copy historical results.
3. Give the corrected analysis a new name and output directory.
4. Run `check`, then annotations and interval validation.
5. Review harmonization and filter audits.
6. Run LDAK and compare declared statistics without inventing a tolerance.
7. Build networks and review node/edge counts.
8. Run the intended HHN mode and validate its mode-specific outputs.

Changing aliases, coordinates, tissue selection, classification, LDAK settings,
network thresholds or interaction backend creates a new profile. It is not a
resume of the old analysis.

## Historical initializers

`initialize_analysis()` currently supports only the historical `v2` and `v3`
templates. Those templates are retained for compatibility and are not the
recommended way to create a corrected E092/Gastric analysis. Use the examples
under `inst/extdata/vignette_examples` as described in
`vignette("analysis-profiles-and-running", package = "hotnetR2")`.
