---
title: "Portable v2 and v3 reproduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portable v2 and v3 reproduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

hotnetR2 combines the hotnetR data interfaces with the v2/v3 regulatory
workflow. Both profiles use all-tissue ENSG harmonization, gastric JEME/HiC,
class-1 enhancers, and Networks 1–2. The v2 target is a gene interval; v3 uses
the historical strand-aware TSS conversion. Compatibility arithmetic is
preserved explicitly rather than silently correcting historical results.

```{r}
project <- tempfile("portable-example-")
config_file <- hotnetR2::initialize_analysis(project, "v3")
config <- hotnetR2::read_analysis_config(config_file)
hotnetR2::resolve_networks(config)
```

Provision large resources separately using `capture_resources()` and
`provision_resources()`. The manifest verifies every file using SHA-256.
Public URLs may be supplied in a manifest, or the user may restore a local
archival bundle. GWAS and LD resources are not included in the R package.

For JEME-only analyses, set `regulatory.hic.tissue_type: []` (or `null`) in the
YAML. HiC is then excluded from both annotations and networks, without requiring
HiC cached files or alias mappings. Start a fresh output directory and regenerate
all stages when changing source selection. Nonempty HiC settings retain the
v2/v3 all-tissue annotation policy; `get_hic(NULL)` remains an all-tissue reader.

```{r eval=FALSE}
hotnetR2::provision_resources("resources.tsv", project,
                            source_root = "/path/to/archival-bundle")
hotnetR2::prepare_ldak_annotations(config, dry_run = FALSE)
hotnetR2::classify_enhancers(config)
hotnetR2::generate_ldak_jobs(config, dry_run = FALSE)
hotnetR2::generate_ldak_local_runner(config, dry_run = FALSE)
# Execute the generated LDAK runner before summarizing results.
hotnetR2::summarize_analysis_ldak_results(config, dry_run = FALSE)
hotnetR2::build_hhotnet_networks(config, dry_run = FALSE)
hotnetR2::generate_hhotnet_job(config, dry_run = FALSE)
# Execute the generated HotNet runner before summarizing clusters.
hotnetR2::summarize_hhotnet_results(config)
hotnetR2::export_hhotnet_graphs(config)
hotnetR2::validate_hhotnet_exports(config)
```

Use `compare_artifacts()` with a declared file inventory. Byte identity of
deterministic outputs is stronger evidence than matching counts. Gzip content
comparison is available explicitly; absolute-path and runtime metadata
differences require separately declared comparisons. Native external-tool
reproducibility across architectures remains distinct from reproduction of
package-owned outputs using frozen raw LDAK/HotNet results.

```{r include=FALSE}
unlink(project, recursive = TRUE)
```
