---
title: "Install hotnetR2 and provision resources"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Install hotnetR2 and provision resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Installation

Install a checked source tarball into the R library used by the analysis
launcher:

```bash
R CMD INSTALL /path/to/hotnetR2_VERSION.tar.gz
Rscript --vanilla -e 'packageVersion("hotnetR2")'
```

For development, install from a reviewed source checkout with standard R
package tools. Do not install from an analysis-local copy of helper functions.

## External software

The R package does not bundle LDAK or Hierarchical HotNet.

| Tool | Required for | Configuration |
|---|---|---|
| LDAK | Annotation cutting and regional association tests | `reference.ldak_exe` |
| PLINK binary reference | LD calculation | `reference.bfile_prefix` |
| bedtools | Interval overlap/classification stages | Available on `PATH` |
| Python | Hierarchical HotNet | `hhotnet.python_executable` |
| Hierarchical HotNet checkout | Similarity, hierarchy and clusters | `hhotnet.local_dir` |
| Apptainer image | Optional container execution | `hhotnet.apptainer_image` |

Validate the exact Python/native Hierarchical HotNet environment independently;
a source checkout without its compiled extension is incomplete.

## Package-managed reference setup

Installation does not download data. From the project root, explicitly create
the cache:

```{r setup-data}
library(hotnetR2)

setup_data(
  cache_dir = ".cache",
  overwrite = FALSE,
  create_harmonization = TRUE,
  type = "lasso"
)
```

This provisions the current package sources for:

- JEME lasso enhancer--target predictions [@Cao2017JEME];
- Jung promoter--other and promoter--promoter data [@Jung2019HiC];
- the NCBI GRCh37.p13 feature table;
- GTEx/GENCODE and HGNC resources used for harmonization.

`overwrite = FALSE` protects an existing cache. Do not refresh a resource in
place for an established analysis; create a new cache or analysis profile and
record the change.

### Alias creation and use

Newly built alias tables carry an explicit raw-to-approved mapping direction.
Unmarked historical caches retain their historical interpretation. The current
recommended workflow uses the curated, direction-marked aliases and records
unmapped, ambiguous and changed identifiers in audit tables.

JEME targets are mapped ENSG-first using all tissues before the selected tissue
is used for network edges. HiC lacks the same target ENSG field and therefore
uses gene symbols and aliases. GENCODE v39 GRCh38 is an identifier resource in
the current GRCh37 analysis, not a promoter-coordinate source.

## STRINGdb setup

STRING acquisition is explicit and separate from `setup_data()`:

```{r string-setup}
download_stringdb_data(
  cache_dir = ".cache/STRINGdb",
  version = "12.0",
  species = 9606,
  score_threshold = 700,
  overwrite = FALSE
)
```

The analysis requires all three version-matched files:

```text
.cache/STRINGdb/9606.protein.aliases.v12.0.txt.gz
.cache/STRINGdb/9606.protein.info.v12.0.txt.gz
.cache/STRINGdb/9606.protein.links.v12.0.txt.gz
```

Analysis execution reads these files offline and never downloads them silently.
The current E092/Gastric profile includes interactions with
`combined_score >= 700`. Only existing promoter-gene nodes are eligible;
enhancers and STRING-only endpoints are excluded. Mapping conflicts use the
documented compatibility rule and are written for review.

## Restore a frozen bundle

For a portable or restricted-data analysis, prefer a checksummed local bundle:

```{r resource-bundle}
manifest <- hotnetR2::capture_resources(
  root = "/path/to/frozen-inputs",
  files = c(
    ".cache/LDAK/GCF_000001405.25_GRCh37.p13_feature_table.txt.gz",
    ".cache/gencode.v39.GRCh38.genes_nodup.rds",
    ".cache/alias_link_nodup.rds"
  ),
  file = "resources.tsv"
)

hotnetR2::provision_resources(
  "resources.tsv", root = "/path/to/new-project",
  source_root = "/path/to/frozen-inputs"
)
```

The manifest establishes byte identity of supplied files. It does not establish
how an undocumented upstream dataset was originally generated.

## Expected cache roles

| Cache | Assembly / role |
|---|---|
| JEME promoter `location` | GRCh37, 1-based TSS metadata |
| JEME enhancer bounds | GRCh37 BED intervals |
| Jung HiC promoter labels | Symbols rematched to NCBI coordinates |
| Jung HiC other-end fragments | GRCh37 BED intervals |
| NCBI feature table | GRCh37 1-based inclusive gene bounds |
| GENCODE v39 | GRCh38 identifier harmonization only |
| STRING v12 | Protein identifiers and association scores; no genomic coordinates |

## Verify before computation

```{r verify}
config <- hotnetR2::read_analysis_config(
  "analysis_E092_Gastric_STRINGdb/analysis.yaml"
)
inputs <- hotnetR2::check_analysis_inputs(config, write_report = FALSE)
print(inputs, row.names = FALSE)
stopifnot(all(inputs$status == "OK"))

hotnetR2::setup_analysis(config, dry_run = TRUE)
```

The preflight is read-only presence and boundary validation. Preserve source
hashes, tool versions and environment information with the run.

## References
