---
title: "Promoter Gene-Symbol Harmonization Report"
author: "Buas analysis"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(knitr)
})

data_dir <- dirname(knitr::current_input(dir = TRUE))
jeme <- read_excel(file.path(data_dir, "final_harmonize_jemeF.xlsx"))
hic_po <- read_excel(file.path(data_dir, "final_harmonize_hiC_PO.xlsx"))
hic_pp <- read_excel(file.path(data_dir, "final_harmonize_hiC_PP_pairs.xlsx"))

mapping_jeme <- jeme |>
  transmute(dataset = "JEME", promoter_side = "promoter", original = id,
            final = symbol, priority)
mapping_po <- hic_po |>
  transmute(dataset = "HiC-PO", promoter_side = "promoter", original = id,
            final = symbol, priority)
mapping_pp <- bind_rows(
  hic_pp |> transmute(dataset = "HiC-PP", promoter_side = "promoter1 (Pro)",
                      original = Pro, final = Pro_sym),
  hic_pp |> transmute(dataset = "HiC-PP", promoter_side = "promoter2 (B)",
                      original = B, final = B_sym)
) |>
  mutate(priority = NA_integer_)

mapping_all <- bind_rows(mapping_jeme, mapping_po, mapping_pp) |>
  mutate(across(c(original, final), as.character)) |>
  filter(!is.na(original), original != "")

mapping_summary <- function(data) {
  data |>
    group_by(dataset, promoter_side) |>
    summarise(
      rows = n(),
      unique_original = n_distinct(original),
      unique_final = n_distinct(final, na.rm = TRUE),
      discordant_rows = sum(!is.na(final) & original != final),
      missing_final = sum(is.na(final) | final == ""),
      .groups = "drop"
    )
}

conflict_table <- function(data) {
  data |>
    distinct(dataset, promoter_side, original, final) |>
    group_by(dataset, promoter_side, original) |>
    summarise(
      n_final = n_distinct(final, na.rm = TRUE),
      final_symbols = paste(sort(unique(final[!is.na(final)])), collapse = "; "),
      .groups = "drop"
    ) |>
    filter(n_final > 1)
}

reverse_conflict_table <- function(data) {
  data |>
    distinct(dataset, promoter_side, original, final) |>
    filter(!is.na(final), final != "") |>
    group_by(dataset, promoter_side, final) |>
    summarise(
      n_original = n_distinct(original),
      original_symbols = paste(sort(unique(original)), collapse = "; "),
      .groups = "drop"
    ) |>
    filter(n_original > 1)
}
```

# Scope and input files

The supplied files represent three different objects:

* `final_harmonize_jemeF.xlsx` is a JEME lookup key. `id` is the raw label and
  `symbol` is the final label.
* `final_harmonize_hiC_PO.xlsx` is a HiC promoter-only lookup key with the same
  `id` and `symbol` convention.
* `final_harmonize_hiC_PP_pairs.xlsx` retains HiC promoter-promoter pair data.
  `Pro` is promoter1 and `B` is promoter2; `Pro_sym` and `B_sym` are their
  final labels.

The report treats repeated tissue or pair observations as observations, not as
mapping conflicts. Conflict checks first reduce each table to distinct
`original`/`final` mappings.

# Basic statistics

```{r basic-statistics}
mapping_summary(mapping_all) |>
  kable(caption = "Rows and discordances by dataset and promoter side")
```

```{r file-statistics}
tibble(
  dataset = c("JEME", "HiC-PO", "HiC-PP"),
  file_rows = c(nrow(jeme), nrow(hic_po), nrow(hic_pp)),
  file_columns = c(ncol(jeme), ncol(hic_po), ncol(hic_pp)),
  distinct_raw_labels = c(n_distinct(jeme$id), n_distinct(hic_po$id),
                         n_distinct(c(hic_pp$Pro, hic_pp$B))),
  distinct_final_labels = c(n_distinct(jeme$symbol), n_distinct(hic_po$symbol),
                            n_distinct(c(hic_pp$Pro_sym, hic_pp$B_sym)))
) |>
  kable(caption = "Workbook-level dimensions")
```

For HiC-PP, the following table makes the promoter-side origin explicit.

```{r hic-pp-sides}
mapping_pp |>
  group_by(promoter_side) |>
  summarise(
    pair_rows = n(),
    unique_raw_labels = n_distinct(original),
    unique_final_labels = n_distinct(final),
    discordant_rows = sum(original != final, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(caption = "HiC-PP promoter1 versus promoter2")
```

# Duplicate and inconsistent mappings

## Exact repeated mappings

These counts describe repeated instances after the source tables are combined.
They are expected for HiC-PP because the same pair can occur in many tissues.

```{r repeated-mappings}
mapping_all |>
  count(dataset, promoter_side, original, final, name = "n_observations") |>
  summarise(
    .by = c(dataset, promoter_side),
    distinct_mappings = n(),
    repeated_mapping_keys = sum(n_observations > 1),
    repeated_observations = sum(pmax(n_observations - 1, 0))
  ) |>
  kable(caption = "Repeated mapping observations")
```

## One original label mapped to multiple final labels

```{r forward-conflicts}
forward_conflicts <- conflict_table(mapping_all)
forward_conflicts |>
  count(dataset, promoter_side, name = "conflicting_original_labels") |>
  kable(caption = "Original labels with more than one final label")
```

```{r forward-conflict-examples}
head(forward_conflicts, 25) |>
  kable(caption = "First 25 forward mapping conflicts")
```

## Multiple original labels mapped to one final label

This is a different kind of many-to-one harmonization and is not necessarily an
error. It is reported because it can change network node counts after labels
are standardized.

```{r reverse-conflicts}
reverse_conflicts <- reverse_conflict_table(mapping_all)
reverse_conflicts |>
  count(dataset, promoter_side, name = "final_labels_with_multiple_originals") |>
  kable(caption = "Final labels receiving multiple original labels")
```

```{r reverse-conflict-examples}
head(reverse_conflicts, 25) |>
  kable(caption = "First 25 reverse mapping collisions")
```

## Cross-dataset consistency

The next table checks whether the same original label appears in multiple
datasets with different final labels. `dataset` and promoter side are retained
so a HiC-PP `Pro`/`B` role is not silently merged with another role.

```{r cross-dataset}
cross_dataset_conflicts <- mapping_all |>
  distinct(dataset, promoter_side, original, final) |>
  group_by(original) |>
  summarise(
    n_dataset_side_mappings = n_distinct(paste(dataset, promoter_side)),
    n_final = n_distinct(final, na.rm = TRUE),
    final_symbols = paste(sort(unique(final[!is.na(final)])), collapse = "; "),
    dataset_sides = paste(sort(unique(paste(dataset, promoter_side))), collapse = "; "),
    .groups = "drop"
  ) |>
  filter(n_final > 1)

tibble(
  conflicting_original_labels = nrow(cross_dataset_conflicts),
  examples = if (nrow(cross_dataset_conflicts) == 0) "None" else
    paste(head(cross_dataset_conflicts$original, 10), collapse = "; ")
) |>
  kable(caption = "Cross-dataset inconsistent original-to-final mappings")
```

# Interpretation

The report distinguishes three quantities that are easy to conflate:

1. A **discordance** is an observation where the raw and final labels differ.
2. A **repeated mapping** is the same raw-to-final pair observed more than once.
3. A **mapping conflict** is one raw label associated with more than one final
   label after exact duplicate mappings are removed.

The JEME and HiC-PO workbooks are keys and should normally have one selected
final symbol per `id`. HiC-PP is pair-level data, so its row count reflects
tissue-specific observations; use the promoter-side mapping tables above when
checking gene-level consistency.
